Model_CFU

### 2016-12-14 Projects/HumanCdGF_cdiff_colonization.Rmd
model tuning mtry/ntree

#### OTUs predictive of CFU (One Mouse)

```{r one mouse model, eval=F}

#One per cage
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    filter(!is.na(log_cfu_med)))
one_mouse_df$human_source <- factor(one_mouse_df$human_source)

om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
om_pred <- c()
om_rsq <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_df %>% 
      group_by(cage_id) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    om_pred <- rbind(om_pred,
                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                          predicted=predict(rf_om, test)))
    om_rsq <- rbind(om_rsq, c(paste(train$mouse_id, collapse = ' '), 
                              tail(rf_om$rsq,1),
                              1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
    om_import_otus <- select(om_import_otus, -IncNodePurity)
}
om_rsq <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))

@@

om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
colnames(om_import_otus) <- seq(ncol(om_import_otus))
om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
om_otu_rank <- om_otu_rank*as.numeric(om_rsq[,3])
om_raw_rank <- sort(apply(om_otu_rank, 1, function(x) sum(x, na.rm = T)), decreasing = T)
med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
                                 importance = om_raw_rank)

>>>>

#### Modeling CFU with One Mouse per Human Donor
In an effort to remove any potential donor community dependency, the model is trained on a single mouse per human donor source, and then tested on all the remaining mice

##### With all OTUs
```{r one mouse model}

#One per source
one_mouse_df <- data.frame(colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    filter(!is.na(log_cfu_med)))
one_mouse_df$human_source <- factor(one_mouse_df$human_source)

om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, -mouse_id, -cage_id, -log_cfu_med)))
om_pred <- c()
om_rsq <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_df %>% 
      group_by(human_source) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
                             importance = TRUE, keep.forest = T, ntree = n_trees)
    om_pred <- rbind(om_pred,
                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
                          predicted=predict(rf_om, test),
                          iter = i))
    om_rsq <- rbind(om_rsq, c(iter = i,
                              mice = paste(train$mouse_id, collapse = ' '), 
                              train_rsq = tail(rf_om$rsq,1),
                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
      select(-IncNodePurity)
}
om_rsq <- data.frame(iter = as.numeric(om_rsq[,1]),
           mice = om_rsq[,2],
           train_rsq = as.numeric(om_rsq[,3]),
           test_rsq = as.numeric(om_rsq[,4]),
           stringsAsFactors = F)
om_rsq_overall <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))

@@

all_otus_importance <- gather(om_import_otus,OTU, import_val,-otu)[,-2]
all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
                      filter(pos_otu, votes == 1) %>% select(Freq) - 
                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
              arrange(desc(Freq))
```

When using all OTUs the model preforms poorly. The median Rsq of the trainset was `r median(om_rsq$train_rsq)`, this is likely do to the features. This is a summary of the importance scores for all predictors:  
`r summary(unlist(select(om_import_otus, -otu)))`  
`r round(100*sum(om_import_otus[,-1]<=0, na.rm = T)/length(om_import_otus[,-1]<=0),2)`% of OTUs have an importance of 0 or less. The median Rsq of the test set was `r median(om_rsq$test_rsq)` and the Rsq of all predicted values was `r om_rsq_overall`.

This one mouse per source model is improved through reducing the amount of negative features, either utilizing the feature set via the top X % of predictors from the model of all mice or take only the features that are more often have an importance > 0 (OTUs : `r as.character(pos_otu$predictor)`).

```{r one mouse model - with OTUs positive in All OTU one mouse model}

#One per source - > 0
one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    filter(!is.na(log_cfu_med))) %>% 
    select(mouse_id, human_source, log_cfu_med,
           one_of(as.character(pos_otu$predictor)))
one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)

om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
om_pred_pos <- c()
om_rsq_pos <- c()
for(i in 1:iters){
    set.seed(i)
    train <- one_mouse_pos_otus_df %>% 
      group_by(human_source) %>% 
      sample_n(1) %>% ungroup() 
    test <- one_mouse_pos_otus_df %>% 
      filter(!mouse_id %in% train$mouse_id)
    for(m_try in 1:25){
      for(n_tree in seq(200, 2000, 100)){
    set.seed(seed)
    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
                             importance = TRUE, ntree = n_tree, mtry = m_try)
    om_pred_pos <- rbind(om_pred_pos,
                     data.frame(select(test, mouse_id, log_cfu_med),
                          predicted=predict(rf_om, test),
                          iter = i,
                          rf_tree = n_tree,
                          rf_try = m_try))
    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
                                      try = m_try,
                                      tree = n_tree,
                              mice = paste0(paste(train$mouse_id, collapse = ' '), '_', m_try, '_', n_tree), 
                              train_rsq = tail(rf_om$rsq,1),
                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
    import_df <- importance(rf_om)
    colnames(import_df)[1] <- paste0(paste(train$mouse_id, collapse = ' '),'_', m_try, '_', n_tree)
    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
      select(-IncNodePurity)
    }}
}
om_rsq_pos_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
                            try = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='try']),
                            tree = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='tree']),
                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
           train_rsq = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
           test_rsq = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
           stringsAsFactors = F)
om_pf_df <- om_rsq_pos_df

om_rsq_pos_overall <- 1 - (sum((om_pred_pos$log_cfu_med - om_pred_pos$predicted)^2)/
               sum((om_pred_pos$log_cfu_med - mean(om_pred_pos$log_cfu_med))^2))

om_pred_pos <- rename(om_pred_pos, observed = log_cfu_med)

all_otus_importance <- gather(om_import_pos_otus,OTU, import_val,-otu)[,-2]
all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
                      filter(pos_otu, votes == 1) %>% select(Freq) - 
                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
              arrange(desc(Freq))

ggplot(om_rsq_pos_df, aes(x = try, y = test_rsq)) + 
  stat_summary(fun.y=median, geom="line", aes(group = iter)) + 
  stat_summary(fun.data = 'median_hilow') +
geom_line() + facet_wrap(~tree, nrow=4)

ggplot(filter(om_rsq_pos_df, try ==6), aes(x = tree, y = test_rsq)) + 
  stat_summary(fun.y=median, geom="line", aes(group = iter)) + 
  stat_summary(fun.data = 'median_hilow')
```

`r median(om_rsq_pos_df$train_rsq)`
`r median(om_rsq_pos_df$test_rsq)`
`r om_rsq_pos_overall`

OM > 0
> median(om_rsq_pos$train_rsq)
[1] 0.08652421
> median(om_rsq_pos$test_rsq)
[1] 0.5331836
> om_rsq_pos_overall
[1] 0.5534567

All mice top 10%
> median(om_rsq_pos$train_rsq)
[1] 0.07336261
> median(om_rsq_pos$test_rsq)
[1] 0.6070359
> om_rsq_pos_overall
[1] 0.633238

> median(om_rsq_pos_df$train_rsq)
[1] 0.05986871
> median(om_rsq_pos_df$test_rsq)
[1] 0.5898076
> `r om_rsq_pos_overall`
> om_rsq_pos_overall
[1] 0.6150207

one mouse - using top 10% of predictive features for all mice in the training set
optimal at
train 
  try = 10, tree = 600
test
  try = 6, tree = 600
but the test set has two groupings in performance. one around a Rsq 0.75, one at 0.5 and two iterations at 0.325

       mouse         high       low          med MeanDecreaseAccuracy MeanDecreaseGini
1      D_507 24.746833562  1.737270 20.102178345          24.73595607        8.4911043
2      C_596 23.285539876  1.001002 20.447905674          24.43000483        7.6690227
3      D_527 18.032458266  1.001002 15.500296401          18.71071088        4.9877513
4      A_566 12.032722269  0.000000 12.067340706          13.81516364        3.0119082
5      D_540 10.866644663  1.737270 12.253700494          13.61280464        2.2541893
6      C_516 11.047770210  1.001002 10.744008798          12.43936737        2.2192996
7      A_503 10.956589615  4.195038  6.774750778          10.79692514        2.2143605
8      A_526  7.778933002  0.000000  6.659309475           8.50216395        1.1996857
9    369_992 -0.875534077  3.932419  4.882427149           3.72993812        1.3625130
10   369_942 -0.596862323  1.417051  3.309012343           2.37657639        0.7644884
11  IN3_2318  1.043691934  2.664468  1.245180833           2.18825289        0.8340440
12  MOUT_896  2.014382262 -2.247333  1.336156704           2.17645115        0.8413007
13   430_920  1.549432277  0.000000  1.108030915           1.91835030        0.6586209
14  MOUT_568  0.666044064 -1.417051  1.909190798           1.75633167        0.7946055
15  IN3_2336 -0.217030290  0.000000  2.406333245           1.45691227        0.5817839
16   NP2_979  1.078030444  0.000000  1.206469792           1.43939415        0.6973072
17   OUTA_NT  0.697539079  1.417051  0.845326576           1.18172290        0.7653337
18   NP2_983 -0.727654094  1.001002  1.931180990           1.05707031        0.8507964
19 OUTA_2063  1.808105182 -1.417051 -0.047912339           0.89013348        0.7126412
20   430_307  0.261169822 -2.464320  1.107393946           0.67386593        0.7143389
21   430_964 -0.574499626  1.001002  1.123257626           0.56350353        0.4823411
22   369_939 -0.619215217  0.000000  1.274820050           0.56227756        0.5328745
23 OUT2_2510  0.009109248 -1.001002  0.721274725           0.37736881        0.6381838
24   OUT2_NT -0.430535085  0.000000  0.460746344           0.14766696        0.6223585
25  IN2_2613  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
26  IN2_2629  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
27  IN2_2673  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
28  NP2_2630  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
29   LINE_NT  1.730524126  0.000000 -1.610672526          -0.09130011        0.6774122
30   430_304  0.153756741 -2.008048 -0.653527605          -0.51088419        0.6561920
31 OUTA_2062 -0.244531023 -1.737270 -0.443149129          -0.70889274        0.7545123
32  LINE_978 -1.613030982  1.001002 -0.003319811          -1.09823532        0.5058066
33   369_331 -2.611303629  0.000000 -0.024601226          -1.37881023        0.6251534
34  LINE_MNT -0.399645816 -1.737270 -1.503302377          -1.75744745        0.5979079
35  LINE_188 -1.752110136 -1.001002 -1.104193541          -1.91010075        0.5634185
36    IN3_NT -2.122086607  0.000000 -1.213308576          -2.28278894        0.7106271
37 OUT2_2533 -1.443012131  0.000000 -2.212914984          -2.28741305        0.5829754

```{r}
om_tf_df$group_mean <- 'med'
om_tf_df$group_mean[om_tf_df$test_rsq > 0.65] <- 'high'
om_tf_df$group_mean[om_tf_df$test_rsq < 0.4] <- 'low'
mice_group <- filter(om_tf_df, try == 5, tree == 500) %>% 
  select(mice, group_mean)
mice_group$mice <- gsub('_\\d*_\\d*$', '', mice_group$mice)

mice_diff <- data.frame(matrix(rep(NA, nrow(mice_group)*length(unique(colonized_cfu$mouse_id))),
                               nrow=nrow(mice_group)))
colnames(mice_diff) <- unique(colonized_cfu$mouse_id)
for(i in 1:nrow(mice_diff)){
    mice_diff[i,] <- ifelse(colnames(mice_diff) %in% strsplit(mice_group$mice, ' ')[[i]], 0,1)
}
predict_rsq <- cbind(mice_group, mice_diff) %>% 
  select(-mice)
predict_rsq$group_mean <- factor(predict_rsq$group_mean)
rsq_grp_rf <- randomForest(predict_rsq[,-1],predict_rsq$group_mean, importance = T)
importance_rsq <- data.frame(mouse = dimnames(importance_rsq)[[1]],importance(rsq_grp_rf))
arrange(importance_rsq, desc(MeanDecreaseAccuracy))
str(importance_rsq)
